Normalization Example using spike-in control as a standard

### 2017-08-07 examples/Example_001_ER_CTCF.Rmd
Writing descriptive text for examples

---
title: "Normalization Example using internal CTCF peaks as a standard"
author: "Andrew Holding"
date: "8/7/2017"
output: pdf_document
---

@@

This is a normalisation using linear regression to CTCF of MCF7 cells chip treameant etc.

@@

These functions facilitate the normalisation of data.

```{r brundle}
source('../package/brundle.R')
```

@@

@@

To keep file size down these are provided as a Rdata File rather than as raw counts.

@@

```{r}
#Load Sample Ids from control sample sheet.
jg.sampleIds <- jg.getSampleIds(jg.controlSampleSheet)
```

@@

## Extract Peak set from DiffBind
jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)
jg.controlPeakset    <- jg.dbaGetPeakset(dbaControl)

@@

@@

@@

##Plot showing normalization calculation (Optional)
  jg.plotNormalization(jg.controlCountsTreated,
                     jg.controlCountsUntreated)
```

```{r}
##Get Normalization Coefficient
jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

@@

@@

```{r}
jg.dba_analysis<-dba.analyze(jg.dba)
dba.plotMA(jg.dba_analysis,bFlip=TRUE)

>>>>

---
title: "Normalization example using internal CTCF peaks as a standard"
author: "Andrew Holding"
date: "8/7/2017"
output: pdf_document
---

@@

This is a worked example of our method to normalize ER ChIP-seq data using CTCF binding to provide an internal control.

## Preprocessing.

To prepare the data for the differential binding analysis the data had to be aligned and CTCF and ER peaks called. As our CTCF control doesn't require any special aligment, data was aligned as for a standard ChIP-seq experiment. Peak calling was done with macs2 using the script shown below.

```{bash, eval=FALSE}

#Calls peaks for all files in a directory against a single control. Change
# <filename> to the filename of you input sample.

 control=<filename>

 for bam in *bam
 do
   root=`basename $bam .bam`
   macs2 callpeak -t $bam -c $control -f BAM -n $root -g hs
 done

As macs2 cannot differentiate between a peak arising from CTCF binding or ER binding we used two samples were squence to establish a CTCF binding consensus. One treated and one untreated. These peaks were then merged as follows.

```{bash, eval=FALSE}

cat  SLX-14438.D701_D504.HKN27BBXX.s_8.r_1.fq.gz.bam_peaks.xls \
      SLX-14438.D702_D503.HKN27BBXX.s_8.r_1.fq.gz.bam_peaks.xls > CTCF_merged.bed

grep -v \# CTCF_merged.bed  | grep -v start > CTCF_merged_filtered.bed
sort -k1,1 -k2,2n CTCF_merged_filtered.bed > CTCF_merged_sorted_filtered.bed

bedtools merge -i CTCF_merged_sorted_filtered.bed  > CTCF_union.bed

We then used this CTCF consensus to establish specfic CTCF or ER binding in the experimental samples. Either taking the interesection of the CTCF consensus and the peak file for a particluar sample to provide a list of CTCF binding sites in the sample, or subratracting the consensus CTCF peaks from the macs2 output to provide a list ER binding sites for that sample.

```{bash, eval=FALSE}

mkdir CTCF
for f in *.narrowPeak
do
 	echo $f
 	bedtools intersect -a $f -b CTCF_union.bed  > CTCF/$f 
done

mkdir ER
for f in *.narrowPeak
do
 	echo $f
 	bedtools subtract -b CTCF_union.bed -a $f -A > ER/$f 
done

These were then used to write two DiffBind samplesheets as described in the DiffBind vignette. One using the CTCF peak list generated from the above scripts, this is the control sample sheet. The second uses the ER peaks, this is the experimetnal sample sheet. These two sample sheets are then provided as the input in the analysis below.

@@

Brundle provies a series of functions we developed for this pipeline that are used to normalizing ChIP-seq data. The usage of these functions are explained in the example that follows.

```{r brundle,results='hide', message=FALSE, warning=FALSE}
source('../package/brundle.R')
```

@@

The intial settings for the analysis are minimal, jg.controlMinOverlap is set to 5 as this is 1 less than the total sample count and we want a strong consensus of CTCF peaks to normalise against. The control and experiment sample sheets are as described above. The conditions reflect the treated and untreaded condition we have used when writing our sample sheet.

@@

Typically we would load the raw data using Diffbind as we are plan to return our normalized data as a Diffbind object.

In this example, as the BAM files which are greater in size than the github filesize limit, we have included the data as the Diffbind object these function would output and instead load that directly. If the code doesn't find the R data object it will try to regenerate if from the raw data.

@@

Diffbind allows us to extract the counts in each peak, we use this function to obtain the data in a form that is easier to normalize.

@@

#  Extract Peak set from DiffBind

jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)
jg.controlPeakset    <- jg.dbaGetPeakset(dbaControl)

@@

To normalize we need to seperate samples out by condition, this is done using the following function. The peakset contains the reads in peaks, the samplesheet provides the metadata needed and the jg.treatedCondition/jg.untreatedCondition specify which condition we want to get the counts for.

@@

At the same time we can extract the samples names for each condition to use later.

@@

The normalization factor is generated by a linear regression generated from the mean number of counts for each peak before and after treatment. The gradient of the linear fit should be equal to 1 given the assumtion that before and after treatment the majority of the control peaks do not change.

This plot is not neaded to caluclate the normalization factor but it does provide a visual interpretation of the process.

```{r, message=FALSE}

# Plot showing normalization calculation (Optional)

jg.plotNormalization(jg.controlCountsTreated,
                     jg.controlCountsUntreated)
```

There are two parts to the normalization coefficent we use in the final correction. The first is the the coefficent to make the levels of CTCF binding before and after treament equal. The second is because DiffBind normalises against the total read count for each sample, we therefore generated a correction factor to control for this.

```{r}
# Get Normalization Coefficient

jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

Once we have these two peices of information we apply both the normalisation coeffecent and the correction factor to our data.

@@

Finally we return the correct values into the DiffBind object. This requires a modificed version Diffbind as included in the original manuscript. It is expected that this will be supported natively by furture versions of Diffind.

@@

We can then visualise the data as normal using DiffBind.

```{r, message=FALSE}
jg.dba_analysis<-dba.analyze(jg.dba)
dba.plotMA(jg.dba_analysis,bFlip=TRUE)

### 2017-08-07 examples/Example_002_ER_CTCF_DeSeq.Rmd
Writing descriptive text for examples

This is a normalisation using only DeSeq Size Factors to CTCF of MCF7 cells chip treameant etc.

@@

```{r brundle, echo=FALSE}
source('../package/brundle.R')
```

@@

## Load control and experimental DiffBind object

To keep file size down these are provided as a Rdata File rather than as raw counts.

@@

Diffbind was simply used as a convient way to extract peak counts. The actual DESeq Pipeline starts simply needs a matrix so we covert from Diffbind and save as csv to provide a starting point for a DESeq pipeline.

@@

# Pipe line starts here
## Load CSV dataset in DESeq format

@@

Note we use the samplesheet here, not needed but neater.

@@

These are not annotated, but other packages exist to do that.
```{r}

@@

@@

#Plot results

>>>>

This is a worked example of our method alternative to normalize ER ChIP-seq data using CTCF binding to provide an internal control. Unlike our previous example this normalisation is using only DESeq Size Factors. This method was found to be less effective but was used for comparision in the supporting publication. This markdown builds on Example 001 and therefore that should be read first. Example 001 also explains the pre-processing steps.

@@

```{r brundle,results='hide', message=FALSE, warning=FALSE}
source('../package/brundle.R')
```

@@

## Load control and experimental counts from Example 001

In this example Diffbind was simply used as a convient way to extract peak counts and provided consitancy for comparision between methods.

To keep file size down counts are provided as a Rdata file rather than as raw data.

@@

The actual DESeq Pipeline starts simply needs a matrix so we covert from Diffbind and save as csv to provide a starting point.

```{r, message=FALSE}

@@

# DESeq Pipeline starting from CSV

## Load CSV dataset in DESeq format

Here we use the CSV files the example code above generated; however, it is possible to generate the input data through a wide variety of methods and we have provided only one.

@@

To normalize the data we use the DESeq2 function to estimate the size factors. This is a very simple method, and it provides some degree of normalization. We also use the original sample sheet to get the different conditions.

```{r,message=FALSE}

@@

We can export the DESeq2 results in CSV format.

@@

## Plots

The Brundle package also contains some convience functions for plotting the output of DESeq2 inplace of DiffBind. These generate the following figures.

@@

### 2017-08-07 examples/Example_003_ER_H2av.Rmd
Writing descriptive text for examples

This is a normalisation using linear regression to CTCF of MCF7 cells chip treameant etc.

@@

To keep file size down these are provided as a Rdata File rather than as raw counts.

@@

## Extract Peak set from DiffBind
jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)
jg.controlPeakset    <- jg.dbaGetPeakset(dbaControl)

>>>>

This is a normalisation using H2av binding within an external spike-in of droshilia chromatin to provide a control for the analysis of ER binding in MCF7 cells after treatment with Fulvestrant.

@@

As before to keep file size down data are provided as a RdatafFile rather than as raw BAM files.

@@

# Extract Peak set from DiffBind
jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)
jg.controlPeakset    <- jg.dbaGetPeakset(dbaControl)
